|

### 2023-07-07 src/manuscript.qmd
doing some adjustments to front half and adding extra models

Those conducting post-mortem analyses of the 2016 Presidential election of Donald J. Trump, a presidential candidate that personifies white grievance politics, contend that the substantial presence of non-college-educated white supporters explains his electoral success [@griffin_et-al_2017]. Analyses by political scientists largely attribute Trump's success in attracting voters to those who see their whiteness as a politically relevant identity where they are distinct from Americans of other racial groups [@jardina_2020_pp]. What remains an open question is whether it actually is whites' concerns about economic status that motivates whites to use their race as a lens to evaluate politics or whether it is a backlash to changes in the racial groups reflected by those occupying visible political offices.

While this is an ongoing debate, part of this disagreement between scholars about where white identity comes from may be an artifact of the traditions of the theoretical frameworks we use in this area. Scholars studying identity, more broadly, often focus on economic threats as a critical motivator for an expressed connection to a social in-group [see @huddy_2001]. However, some scholars studying white political attitudes and racial identity tend to focus on the role of political threats [see @bartels_2020]. This mismatch leads to some conceptualizing white identity as originating from economic views in the tradition of the identity literature. In contrast, others conceptualize white identity as motivated by political threats.

Along with providing a narrative that scholars might use to identify potential mechanisms connecting political threats to white identity, I examine the correlates of white identity. Using data from the 2012, 2016, and 2020 American National Election Study, I utilize penalized regression to test the claim that a handful of predictors that reflect attitudes about political power are more influential than general (sociotropic) and pocketbook (egotropic) attitudes about the economy. I find that proxies of political threat appear stronger than economic attitudes in the post-Trump era. As this particular analysis is not a causal test of the mechanism connecting threat to white identity, I close by arguing that scholars should continue to examine the mechanisms that explain this connection.

@@

Social identity theory suggests that individuals form attitudes based on groups they associate with, even if those groups are subjectively defined [see @tajfel_turner_1986]. In the famous Robber's Cave experiment, @sherif_1988 divided a group of adolescent boys into two groups arbitrarily. After a week of spending time with others in the group, the boys came up with group names that defined the two groups and distinguished themselves; one group called themselves the Eagles, and the other group referred to themselves as the Rattlers. With the introduction of competition, these boys reported fierce hostility toward the members of the other group and expressed positive attitudes toward those within the group. 
<!--
This example illustrates several features of social identities. The first feature of social identity theory is that these groups we may associate ourselves with are often subjectively defined [see @huddy_2001]. The second feature is that these subjective definitions of the group make distinctions between the two groups that come to matter to members of each group - some are part of our group, and some are part of that other group. Many cognitive theories suggest that this occurs as a result of the tendency for humans to reduce cognitive load by grouping things together [@huddy_2001]. The third feature is that introducing competition between these groups shapes attitudes toward those within and without your group. The motivator is that groups with higher status often want to keep that status, and individuals in the high-status group are much more likely to support inequality to satisfy that motivation [@burns_kinder_2012;@morrison_et-al_2009]. Fourth, individuals firmly hold these attitudes, which tend to be long-lasting. An additional feature of theories of social identity is that the degree to which one might identify with one group or another is fluid and that whether one identity that one might hold is active or if it is another often depends on the context and the identity's relevance to that context [@huddy_2012].

Racial groups in the U.S. contain a number of these features. As a result, a burgeoning literature in political science that has made this connection takes the racial identity of whites more seriously as one that informs their political attitudes [see @jardina_2019_cup]. Race in America is subjectively defined [see @rile-hayward_2013], and these definitions are deeply embedded and reinforced by our institutions [@lopez_2006]. Whiteness and one's willingness to identify with that group is important [@wong_cho_2005] as it reflects a desire for an association with the history and affordances of being white [@mcdermott_samson_2005]. 
-->
Recent work makes the case that we should seriously consider whites' racial identity as an identity [@jardina_2019_cup]. As an identity, much of the literature view these attitudes resulting from whites' racial identity as stemming from economic threat. Some studying white political identity find economic predictors as substantively and statistically significant [see @perez_et-al_2022_pp;@lindsay_2023_apr on their racial attitudes writ large]. Observational evidence supports this claim as well. For example, Trump's messaging about job market competition with immigrants appeared to be successful at attracting white voters [@bunyasi_2019]. After all, the identity literature, more broadly, often focuses on economic threats. Evolutionary psychologists often take the position that identity and the resulting attitudes and behaviors directed toward out-groups are the result of competition over material resources [see @huddy_2001]. Many political scientists use this as an explanation for negative attitudes directed toward immigrants among those who feel insecure about their job prospects [@pardos-prado_xena_2019]. Others argue that deindustrialization and job loss activate whites' concerns about the racial hierarchy and predict their support for Republican challengers that express support for keeping the racial status quo [@baccini_weymouth_2021_apsr].

@@

Not one single paper or effort to provide empirical evidence can satisfactorily address such a large and complex question. The causal mechanism here is extremely difficult to pin down. In an effort to address this question in a bite-sized chunk, I first start with examining whether or not correlational evidence aligns with my claim that individual-level economic predictors are weaker than perceptions of political influence. In the next section, I examine whether attitudes expressing concern with economic or political threats are more robust correlates of the degree to which someone feels that being white is important to their identity. The goal is not to test particular causal mechanisms but to examine whether my belief and others' that white identity reflects reactions to political threats more than it does economic.

# Data

To examine whether whites' attitudes expressing concern about political threats are more robust correlates than their perceptions of the economy, I use the 2012, 2016, and 2020 American National Election Studies (ANES). The studies are nationally representative questions that capture several questions that lend themselves to testing the relationship between economic and political concerns while also asking white respondents to report how important being white is to their identity.
<!--
Of course, I cannot *prove* that these questions reflect the two types of threats I purport to represent. This leaves any evidence of the particular mechanism connecting these predictors to reported levels of white identity -- among other reasons -- suspect. However, the goal of this paper is not necessarily to test a particular causal mechanism -- this is a much larger project that we should continue to work on collectively. Nevertheless, the paper aims to offer an argument that contextualizes white identity in history and how that leads to different predictions about the correlates of white identity than many might assume from understanding the literature on the origins of social identity.
-->
Each study asks white respondents, "How important is being white to your Identity" and offers them a five-point Likert scale with the options reflecting their level of agreement with the question.^[Details about question-wording, original ANES coding, and my coding for all variables I include in my models are included in @tbl-anes-measures located in the Supplementary Information. I include descriptive statistics for the measures in these samples in @tbl-2012-descriptive-stats, @tbl-2016-descriptive-stats, and @tbl-2020-descriptive-stats located in the Supplementary Information.] They do not ask white respondents to report on items from the white political identity measures of @jardina_2019_cup or the more complex battery of items that @helms_carter_1990 discusses. However, this question does capture the degree to which one sees their whiteness as part of who they are and reflects the operationalization that others have used [see @schildkraut_2017;@jardina_2019_cup;@perez_et-al_2022_pp].

The ANES does not ask questions that explicitly capture the attitudes of respondents about perceptions of the economy for different groups. That is, there are no questions asking respondents about their perceptions of the economy for Black Americans relative to white Americans. The literature suggests, however, that those who have lower incomes are more likely to feel insecure about their financial status and are more likely to use this economic threat as a motivation for their negative attitudes toward immigrants [@pardos-prado_xena_2019]. Therefore, one measure of economic threat that I use is the respondent's reported level of total family income. Those with lower levels of education are also likely to feel similar levels of insecurity about their economic status, as their ability to seek higher-paying jobs is likely more limited. Whites who report that they feel that their family is doing worse relative to the previous year and knowing others who lost a job also reflect a degree of insecurity about their economic status and may reflect perceptions of economic threat [@jardina_2019_cup]. I include an item reflecting respondents' perceptions of economic mobility today relative to the past. In 2012, whites harboring negative attitudes toward Obama rooted in racial animous were more likely to express more negative views of the economy relative to how it had been in the past [@sides_et-al_2018]. Others demonstrate that those who are reacting to economic threats and are concerned about their and other whites' status evaluate the economy and their economic position as worse [@baccini_weymouth_2021_apsr]. Therefore, I additionally include questions about retrospective sociotropic evaluations of the economy.

The ANES includes several questions that reflect perceptions of whites losing political power. First, I include a question that asks respondents to report whether or not they felt that whites currently have too little or too much influence on politics today. I also want to examine the degree to which respondents feel that the institutions benefit non-whites rather than whites by including the question that whites face discrimination and that the government is biased against whites. I include a question asking respondents about their belief that traditions in this country are breaking.^[The ANES only included this question in the 2012 and 2016 questionnaires.] While these, like the economic attitudes, are just proxies of threat, they reflect the extent to which respondents may feel insecure about the political influence of whites.

Furthermore, the ANES includes the racial resentment battery, which asks questions reflecting whites' beliefs that Blacks are individually responsible for their circumstances and undeserving of assistance. I control for racial resentment as the individual items likely confound the relationship between both economic and political threat measures on expressed white identity. As economic evaluations often reflect partisans reporting a better economy when the government is run by co-partisans and a worse economy when the other party runs the government [@achen_bartels_2017], I additionally control for a 7-point item measure of partisanship. Finally, I control for whether or not the respondent is female, as Sex affects one's socioeconomic status.

@@

The summary of the literature points to a large number of potential predictors that influence the degree to which one takes on white political identity. To adjudicate between hypotheses, there is value in a comprehensive model that considers the plausibility of each. Most models consider those predictors they believe are most plausible and do not include others that the researcher believes to be less plausible [see @clarke_primo_2012]. They understandably do this out of concern with increases in the inefficiency of one's regression estimates: the more predictors you include, the risk of overfitting your model to your particular sample and from multicollinearity increases, thereby erroneously generalizing conclusions drawn from the fitted statistical model. Though this concern is reasonable, such a model neglects to examine the effects of the predictors a researcher believes to be less plausible.

Therefore, I use a penalized regression model to balance these concerns. The literature surveyed above suggests that there are a small number of variables that do predict white political identity, and there are many others that have no effect. Imagining an assumed hypothesis about the distribution of coefficients, this implies that the L2 regularization parameter equals 0. This is because the L2 regularization parameter shrinks coefficients asymptotically toward zero. The L1 regularization parameter, however, does allow for a shrinkage parameter to pull coefficients to 0. Given this, it suggests that the more appropriate model to capture this debate would be the LASSO penalized regression.

@@

lasso[['ordinal_tidy']] <- rbind(
  lasso[['2012_ordinal_tidy']]
  , lasso[['2016_ordinal_tidy']]
  , lasso[['2020_ordinal_tidy']]
) |> # append the three models' results together
  mutate(
    term = case_when(
      term == 'b_rboff' ~ 'Retrospective better off'
      , term == 'b_epast' ~ 'Economy better'
      , term == 'b_unpast' ~ 'Unemployment better'
      , term == 'b_edu' ~ 'Education'
      , term == 'b_income' ~ 'Income'
      , term == 'b_losejob' ~ 'Worried about losing job'
      , term == 'b_ecfamily' ~ 'Worried about family finances'
      , term == 'b_ecjob' ~ 'Know someone lost job'
      , term == 'b_tradbreak' ~ 'New lifestyles breaking society'
      , term == 'b_govbias' ~ 'Gov. biased against whites'
      , term == 'b_inflwhite' ~ 'whites influence politics'
      , term == 'b_whitediscrim' ~ 'whites are discriminated against'
      , term == 'b_raceResent' ~ 'Racial resentment'
      , term == 'b_pid' ~ 'Party ID'
      , term == 'b_female' ~ 'Female'
    ), # rename the parameters
    term = factor(
      term
      , levels = c(
        'Retrospective better off'
        , 'Economy better'
        , 'Unemployment better'
        , 'Education'
        , 'Income'
        , 'Worried about losing job'
        , 'Worried about family finances'
        , 'Know someone lost job'
        , 'New lifestyles breaking society'
        , 'Gov. biased against whites'
        , 'whites influence politics'
        , 'whites are discriminated against'
        , 'Racial resentment'
        , 'Party ID'
        , 'Female'
      )
    ) # convert it to a factor variable
  ) |> 
  filter(!is.na(term)) # get rid of na rows
    #** make a data.frame of results for the econ predictors
lasso[["ordinal_tidy_econ"]] <- lasso[["ordinal_tidy"]] |>
  filter(
    term == "Retrospective better off" | term == "Economy better" | 
    term == "Unemployment better" | term == "Education" | term == "Income" | 
    term == "Worried about losing job" | term == "Wrroeid about family finances" |
    term == "Know someone lost job"
  )
    #** make a data.frame of results for the political predictors
lasso[["ordinal_tidy_poli"]] <- lasso[["ordinal_tidy"]] |>
  filter(
    term == "New lifestyles breaking society" | term == "Gov. biased against whites" |
    term == "whites influence politics" | term == "whites are discriminated against" |
    term == "Racial resentment"
  )
  #* plot for econ predictors
econ <- ggplot(
  data = lasso[['ordinal_tidy_econ']]
  , aes(
    y = term
    , x = estimate
    , xmin = conf.low
    , xmax = conf.high
    , linetype = factor(model)
    , shape = factor(model)
  )
) +
  geom_pointinterval(
    position = position_dodge(width = 0.4)
    , size = 3
) +
  geom_vline(
    xintercept = 0
    , linetype = 2
) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(
    linetype = 'Model'
    , shape = 'Model'
    , x = 'Estimate'
    , y = ''
    , caption = 'Data Source: 2012, 2016, 2020 ANES.\n Dots are median draw from posterior distribution.\n Bars represent 95% credible intervals.'
) + 
  theme(text=element_text(size=12, family = 'sans'))
  #* plot for political predictors
poli <- ggplot(
  data = lasso[['ordinal_tidy_poli']]
  , aes(
    y = term
    , x = estimate
    , xmin = conf.low
    , xmax = conf.high
    , linetype = factor(model)
    , shape = factor(model)
  )
) +
  geom_pointinterval(
    position = position_dodge(width = 0.4)
    , size = 3
) +
  geom_vline(
    xintercept = 0
    , linetype = 2
) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(
    linetype = 'Model'
    , shape = 'Model'
    , x = 'Estimate'
    , y = ''
    , caption = 'Data Source: 2012, 2016, 2020 ANES.\n Dots are median draw from posterior distribution.\n Bars represent 95% credible intervals.'
) + 
  theme(text=element_text(size=12, family = 'sans'))

@@

Examining the figure, it is readily apparent that in the face of indicators of attitudes about political status, economic attitudes inconsistently associate with white political identity. In all three years, reported attitudes about the nation's economy being better relative to the previous year, reported feelings about the nation's unemployment rate, income, concern about one's family finances, and knowing someone who lost a job are not meaningfully correlated with white political identity. For other indicators of economic threat, none appear to be consistently meaningful. Retrospective evaluations of one's financial situation and education matter in 2012, but these effects were not present in 2016 and 2020.

Turning to indicators of political threat, we still see significant inconsistency. In both 2012 and 2016, whether someone feels that new lifestyles are breaking society positively affects reported levels of white political identity. This relationship is meaningful in 2012 and 2016 (this question is not present in the 2020 questionnaire). In 2020, those who reported that the government favored Black Americans also reported a higher white political identity, and those who feel that whites are discriminated against; however, neither of these effects are meaningful in 2012 or 2016.

@@

Overall, these results suggest that the effects of economic threat on white political identity are more robust in the pre-Trump era. Again, this is not confirmatory as there is no test of a causal mechanism for either economic or political threats. The goal of this analysis is to examine whether, in an omnibus test, either of these threats appears to be a more plausible predictor of white political identity. Though the relationship between political threats is inconsistent, their relationships appear more likely; particularly when we consider how many of these attitudes are explained by a rather complex data-generating process.

Economic threats tend to play a more significant part in explaining white identification in 2012. Political threats, however, appear to be a stronger predictor of white identification in the post-Trump era. In 2012, the U.S. economy was recovering from a massive recession and housing market collapse. However, as Obama's administration continued, conservatives became more outspoken about Obama's race [@tesler_2016]. This shift to focus on Obama's race continued through the 2016 Presidential election and may reflect a much more explicit message to whites about political threats [@sides_et-al_2018]. Though in 2020, the global economy faced significant uncertainty due to the COVID-19 pandemic, economic factors appear to be relatively less important. These results suggest that white political identity results from perceptions of a political threat in the post-Trump era.

@@

## Model results {.appendix}

>>>>

Are White identifiers those who are percieve economic threats or is it a backlash to a loss of influence over politics? The existing work offers an unclear answer to this question. This note attempts to provide an ombnibus test to determine not only which of these questions has more empirical support but also which among common proxies of economic and political power in the United States appear to be reliable predictors of a White individual's expression of racial identity.

A common source of data and of measures for researchers in the United States is the American National Election Study. I use a penalized regression using data from the 2012, 2016, and 2020 American National Election Studies to examine the degree to which individual measures of economic and political influence explain an individual's uptake of White identity as well as to examine the degree to which changes in the political context may increase or decrease the explanatory power of these measures. I find that proxies reporting concerns about Whites' political influence are stronger explantory factors than proxies of economic influence in the post-Trump era.

@@

One perspective on what explains racial identity among Whites in the United States are that they are responding to threats to their economic status and power. Realistic Conflict Theory suggests that a racial identity and racial prejudice come as a response to a collective concern among members of a particular racial group about the loss of economic influence and status a group holds relative to other racial groups [@blumer_1958_psr;@bobo_hutchings_1996_asr]. Some work in support of this perspective sees White racial identity as a response to concerns about increased job market competition through immigration [@pardos-prado_xena_2019]. Other work finds support for this argument in lab experiments [@perez_et-al_2022_pp] and in surveys [@lindsay_2023_apr]. In one of the most thorough treatments on the use of racial identity by Whites in politics, @jardina_2019_cup includes indicators of economic status and power in the primary model to predict White political identity. If this perspective holds, we might expect that in times of economic hardship, Whites may be most sensitive to economic threats as the collective and the individual see economics as most salient to them.

Social identity theory, and its offshoots, would suggest that Whites expressing a racial identity are responding to political, cultural, and societal threats. While these threats are a bit more hard to define, there are a number of examples of Whites expressing prejudice towards other racial groups in response to these sorts of challenges to their power. @tesler_2016 demonstrates that Whites heavily racialized American politics in response to Barack Obama's ascention to the White House. Other work demonstrates that the saliency and reprocussions of aligning with competing racial groups drove non-White political candidates to distance themselves from their racial confederates in an attempt to appease Whites [@stephens-dougan_2020]. Some pundits on Fox News suggest that Whites are in direct competition for jobs with immigrants from South and Central America. Despite these claims, some evidence suggests that Whites' hostility directed toward non-White immigrant groups depends on the degree to which they assimilate [@green_staerkle_2013; see also @hainmueller_hopkins_2014]. In terms of attitudes oriented toward the in-group, @dawson_1994_pup argues that Black Americans hold a strong racial identity that supercedes their class-based identity due to the persistent focus on Black individual's race in their personal and political life. For Whites, their racial identity has long been central to their behavior and the way in which they mobilize to shape legal and political institutions to maintain power [@rile-hayward_2013].

We may also see that economic or political threat may appear differently for partisans and for female respondents. Recent evidence suggests that racial attitudes are guided by partisanship [@engelhardt_2021_bjps]. Specifically, some evidence suggests that White Democrats are much more sensitive to their racial identity than White Republicans due to the higher amount of racial heterogeneity within the Democratic party. This may mean for White Democrats that they have more exposure to intergroup competition for power within the party (such as the choice to support Hillary Clinton versus Barack Obama) and may have higher White identification as a result. The second way that these threats may matter more or less for Whites is on the dimension of sex. Gender-based differences in economic influence and status remains quite high. We therefore may expect that female individuals are more sensitive to economic threats than their male counterparts. In addition to the ombnibus model, I also include two additional models where I split the samples along the dimension of partisanship (including leaners in each) and on the dimension of sex.

# Political versus economic threats

Identity is largely an elusive concept to operationalize. Though there are implicit measures of White identity [see @knowles_peng_2005_jpsp], the common and measure of white identity is a response to the question, "How important is being White to your identity?" This measure captures a significant feature of an identity: the centrality or the importance of that identity to an individual. Though a rather rough measure of something as complex as identity, it has been shown to produce similar results to more complex measures of white identity and holds a significant amount of face validity [@jardina_2019_cup].

Threat is also quite difficult to measure directly. However, we can make a number of reasonable guesses about who may feel economic and political threat. There are two key things that threats affect. Status and power. Status refers to the respectability of the group whereas power refers to the influence that the group has [@craig_phillips_2023_ohpp]. Threat here refers to either a manifestation of economic or political challenges to the current racial hierarchy. I do not distinguish between these two sources of threat as common survey measures are too crude to distinguish between how the threat manifests. Here, I am focusing on what types of non-material threats activate White identity in the context of politics and will, as a shorthand refer to political, cultural, and societal threats as political threats.

A reasonable approach to measuring economic threat is to measure the economic circumstances that White individuals find themselves in. We may expect that Whites who are facing bottom-up economic competition, will be those that experience the strongest desire to identify themselves with their racial group. As the dominant racial group in the United States, Whites that are unemployed, are surrounded by other Whites who are experiencing economic hardship and those who have limited capacity to switch careers are those that may be the most motivated to identify with the dominant racial group to defend Whites' influence and status in society [@bobo_hutchings_1996_asr]. Therefore, measures of employment, finances, family finances, finances of those they know, and educational attainment all seem likely predictors of racial identity for Whites. This fits with measures that @dawson_1994_pup used to measure the influence of economics in Black American's identity. These measures also fit with @lindsay_2023_apr who found that many of these items predicted racial resentment among Whites.

Here I conceptualize of political threat as perceptions that politics and political institutions no longer serves White people. Existing work demonstrates that Whites who express a desire to maintain the racial status quo with Whites on top are more likely to express that Whites are discriminated against at similar levels as Blacks [@marshburn_et-al_2022_gpir]. As Whiteness is often a central component of Whites' conceptualization of American society [see @green_staerkle_2013], changes to "traditional American values" may reflect concern about changes to the influences of White values. Finally, a common finding is that expression of White identity is heavily wrapped up in the perception that racial stratifications are by meritocratic means [@lowery_et-al_2012_jpsp]. As members of a dominant social group, when given information that either threatens one's belief that their merits are earned or that their group operates as a way to maintain inequality, we expect lower levels of expression in identification with that group. The classic items of racial resentment construct capture these ideas in that they revolve around the idea that Blacks are not disadvantaged as the result of systematic oppression by Whites. Though many have often used racial resentment as a measure of prejudice which would be a possible outcome of White identity, some make the compelling argument that it is distinct as a concept and operationalization [@davis_wilson_2022_ucp].

<!--
Social identity theory suggests that individuals form attitudes based on groups they associate with, even if those groups are subjectively defined [see @tajfel_turner_1986]. In the famous Robber's Cave experiment, @sherif_1988 divided a group of adolescent boys into two groups arbitrarily. After a week of spending time with others in the group, the boys came up with group names that defined the two groups and distinguished themselves; one group called themselves the Eagles, and the other group referred to themselves as the Rattlers. With the introduction of competition, these boys reported fierce hostility toward the members of the other group and expressed positive attitudes toward those within the group. 
<!--
This example illustrates several features of social identities. The first feature of social identity theory is that these groups we may associate ourselves with are often subjectively defined [see @huddy_2001]. The second feature is that these subjective definitions of the group make distinctions between the two groups that come to matter to members of each group - some are part of our group, and some are part of that other group. Many cognitive theories suggest that this occurs as a result of the tendency for humans to reduce cognitive load by grouping things together [@huddy_2001]. The third feature is that introducing competition between these groups shapes attitudes toward those within and without your group. The motivator is that groups with higher status often want to keep that status, and individuals in the high-status group are much more likely to support inequality to satisfy that motivation [@burns_kinder_2012;@morrison_et-al_2009]. Fourth, individuals firmly hold these attitudes, which tend to be long-lasting. An additional feature of theories of social identity is that the degree to which one might identify with one group or another is fluid and that whether one identity that one might hold is active or if it is another often depends on the context and the identity's relevance to that context [@huddy_2012].

Racial groups in the U.S. contain a number of these features. As a result, a burgeoning literature in political science that has made this connection takes the racial identity of whites more seriously as one that informs their political attitudes [see @jardina_2019_cup]. Race in America is subjectively defined [see @rile-hayward_2013], and these definitions are deeply embedded and reinforced by our institutions [@lopez_2006]. Whiteness and one's willingness to identify with that group is important [@wong_cho_2005] as it reflects a desire for an association with the history and affordances of being white [@mcdermott_samson_2005]. 
->
Recent work makes the case that we should seriously consider whites' racial identity as an identity [@jardina_2019_cup]. As an identity, much of the literature view these attitudes resulting from whites' racial identity as stemming from economic threat. Some studying white political identity find economic predictors as substantively and statistically significant [see @perez_et-al_2022_pp;@lindsay_2023_apr on their racial attitudes writ large]. Observational evidence supports this claim as well. For example, Trump's messaging about job market competition with immigrants appeared to be successful at attracting white voters [@bunyasi_2019]. After all, the identity literature, more broadly, often focuses on economic threats. Evolutionary psychologists often take the position that identity and the resulting attitudes and behaviors directed toward out-groups are the result of competition over material resources [see @huddy_2001]. Many political scientists use this as an explanation for negative attitudes directed toward immigrants among those who feel insecure about their job prospects [@pardos-prado_xena_2019]. Others argue that deindustrialization and job loss activate whites' concerns about the racial hierarchy and predict their support for Republican challengers that express support for keeping the racial status quo [@baccini_weymouth_2021_apsr].

@@

Not one single paper or effort to provide empirical evidence can satisfactorily address such a large and complex question. The causal mechanism here is extremely difficult to pin down. In an effort to address this question in a bite-sized chunk, I first start with examining whether or not correlational evidence aligns with my claim that individual-level economic predictors are weaker than perceptions of political influence. In the next section, I examine whether attitudes expressing concern with economic or political threats are more robust correlates of the degree to which someone feels that being white is important to their identity. The goal is not to test particular causal mechanisms but to examine whether my belief and others' that white identity reflects reactions to political threats more than it does economic.
-->
# Data

To examine whether whites' attitudes expressing concern about political threats are more robust correlates than their perceptions of the economy, I use the 2012, 2016, and 2020 American National Election Studies (ANES). The studies are nationally representative questions that capture several questions that lend themselves to testing the relationship between economic and political concerns while also asking white respondents to report how important being white is to their identity.

Though I am using the more crude measure of White identity and am limited in the predictors I can choose from, a larger concern is that the degree to which these threats may vary as a result of changes to the political landscape. Specifically, in 2008 the United States experienced a massive recession that lead hit the housing market quite hard. In 2012, the economy was still recovering. In 2016, Donald Trump became infamous for an election where he made race and gender extremely salient [@banda_cassese_2022_pb]. Finally, in 2020, Trump had been in office for four years with an improving economy. Each of these three years reflect a different context and may offer the opportunity to observe how economic and political threats operate differrently.

Question wording for each of the items in the following models are included in @tbl-anes-measures which is located in the Supplementary Materials. I include descriptive statistics for the measures in these samples in @tbl-2012-descriptive-stats, @tbl-2016-descriptive-stats, and @tbl-2020-descriptive-stats located in the Supplementary Information.

@@

The primary contribution of this paper is to examine which of a number of competing measures of different types of threat may predict white identity for individuals. As this requires an ombnibus test, I elect to use a penalized regression. I provide a more detailed discussion of the reasons why I use an estimator with an L2 regularization parameter, often refered to as a LASSO model in the supplementary materials. For a brief comment as to why I use this approach is because approach is designed to take models with a large number of predictors and to shrink the coefficients toward zero. In practice, this often means that predictors that has relatively fewer explanatory power have much smaller coefficients than the variables that do provide significant predictive power.

@@

lasso[['ordinal_tidy']] <- rbind(
  lasso[['2012_ordinal_tidy']]
  , lasso[['2016_ordinal_tidy']]
  , lasso[['2020_ordinal_tidy']]
) |> # append the three models' results together
  mutate(
    term = case_when(
      term == 'b_rboff' ~ 'Retrospective better off'
      , term == 'b_epast' ~ 'Economy better'
      , term == 'b_unpast' ~ 'Unemployment better'
      , term == 'b_edu' ~ 'Education'
      , term == 'b_income' ~ 'Income'
      , term == 'b_losejob' ~ 'Worried about losing job'
      , term == 'b_ecfamily' ~ 'Worried about family finances'
      , term == 'b_ecjob' ~ 'Know someone lost job'
      , term == 'b_tradbreak' ~ 'New lifestyles breaking society'
      , term == 'b_govbias' ~ 'Gov. biased against whites'
      , term == 'b_inflwhite' ~ 'whites influence politics'
      , term == 'b_whitediscrim' ~ 'whites are discriminated against'
      , term == 'b_raceResent' ~ 'Racial resentment'
      , term == 'b_pid' ~ 'Party ID'
      , term == 'b_female' ~ 'Female'
    ), # rename the parameters
    term = factor(
      term
      , levels = c(
        'Retrospective better off'
        , 'Economy better'
        , 'Unemployment better'
        , 'Education'
        , 'Income'
        , 'Worried about losing job'
        , 'Worried about family finances'
        , 'Know someone lost job'
        , 'New lifestyles breaking society'
        , 'Gov. biased against whites'
        , 'whites influence politics'
        , 'whites are discriminated against'
        , 'Racial resentment'
        , 'Party ID'
        , 'Female'
      )
    ) # convert it to a factor variable
  ) |> 
  filter(!is.na(term)) # get rid of na rows
    #** make a data.frame of results for the econ predictors
lasso[["ordinal_tidy_econ"]] <- lasso[["ordinal_tidy"]] |>
  filter(
    term == "Retrospective better off" | term == "Economy better" | 
    term == "Unemployment better" | term == "Education" | term == "Income" | 
    term == "Worried about losing job" | term == "Wrroeid about family finances" |
    term == "Know someone lost job"
  )
    #** make a data.frame of results for the political predictors
lasso[["ordinal_tidy_poli"]] <- lasso[["ordinal_tidy"]] |>
  filter(
    term == "New lifestyles breaking society" | term == "Gov. biased against whites" |
    term == "whites influence politics" | term == "whites are discriminated against" |
    term == "Racial resentment"
  )
  #* plot for econ predictors
econ <- ggplot(
  data = lasso[['ordinal_tidy_econ']]
  , aes(
    y = term
    , x = estimate
    , xmin = conf.low
    , xmax = conf.high
    , shape = factor(model)
  )
) +
  geom_pointinterval(
    position = position_dodge(width = 0.6)
    , size = 4
) +
  geom_vline(
    xintercept = 0
    , linetype = 2
) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(
    shape = 'Model'
    , x = 'Estimate'
    , y = ''
    , caption = 'Data Source: 2012, 2016, 2020 ANES.\n Dots are median draw from posterior distribution.\n Bars represent 95% credible intervals.'
) + 
  theme(text=element_text(size=12, family = 'sans'))
  #* plot for political predictors
poli <- ggplot(
  data = lasso[['ordinal_tidy_poli']]
  , aes(
    y = term
    , x = estimate
    , xmin = conf.low
    , xmax = conf.high
    , shape = factor(model)
  )
) +
  geom_pointinterval(
    position = position_dodge(width = 0.6)
    , size = 4
) +
  geom_vline(
    xintercept = 0
    , linetype = 2
) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(
    shape = 'Model'
    , x = 'Estimate'
    , y = ''
    , caption = 'Data Source: 2012, 2016, 2020 ANES.\n Dots are median draw from posterior distribution.\n Bars represent 95% credible intervals.'
) + 
  theme(text=element_text(size=12, family = 'sans'))

@@

Examining the figure, we see that none of the proxies for economic threat consistently correlate with White political identity; however there is at least one proxy of political threat that does. For economic threats, we see that when the economy was recovering from the recession and Obama was in Office, those who perceived that the economy was better off retrospectively and with a lower reported level of education were correlated with expressing a higher White identity. However, when the economy was better and Trump was in office, we see that those less worried about losing a job were reported a higher white identity.

For political threats, we see that in both instances where Obama was in the White House that those who reported being concerned with "new lifestyles" as well as those with a higher score on the racial resentment measures were associated with having higher white political identities. When Trump was in the White House, we see that those who felt that the government was biased against Whites, that Whites were discriminated against, and who scored higher on the racial resentment measures all reported higher levels of White political identity.

@@

Overall, these results suggest that common proxies of economic threat are at least inconsistent measures of White identity. Even in a time where the economy is recovering from a collapse, more proxies of political threat seemed to predict White identity. When a White president well-known for his penchance for racial dog-whistles takes office, we see support for the idea that White grievance politics was at a high -- high those who felt that political institutions and that the direction of society was turning away from the racial status quo were those that reported their racial identity as more central to who they are.

There are a few important points that are illustrated here. First, suggestions that White political identity is something occuring for those who are most economically threatened appear to have weak theoretical and empirical backing when directly pitted against other forms of threat. Second, that measures of cultural or societal threat are important variables to include if one is using white identity as a variable in a model.

@@

## Model{.appendix}

The literature surveyed above suggests that there are a small number of variables that do predict white political identity, and there are many others that have no effect. Imagining an assumed hypothesis about the distribution of coefficients, this implies that the L2 regularization parameter equals 0. This is because the L2 regularization parameter shrinks coefficients asymptotically toward zero. The L1 regularization parameter, however, does allow for a shrinkage parameter to pull coefficients to 0. Given this, it suggests that the more appropriate model to capture this debate would be the LASSO penalized regression.

### Full tables of results {.appendix}
